---
title: "Quantium Virtual Internship - Retail Strategy and Analytics - Task 1"
author: "Wayan Gracias"
date: "2023-03-07"
output: pdf_document
---

#### Load required libraries

```{r}
library(tidyverse)
library(data.table)
library(readxl)
library(ggmosaic)
library(tidyr)
```

#### Load datasets

```{r}
customerData <- read.csv("QVI_purchase_behaviour.csv")
transactionData <- read_xlsx("QVI_transaction_data.xlsx")
```

## Exploratory Data Analysis

The first step in any analysis is to first understand the data. Let's take a look at each of the datasets provided.

### Examining transaction data

We can use `str()` to look at the format of each column and see a sample of the data. As we have read in the dataset as a `data.table` object, we can also run `transactionData` in the console to see a sample of the data or use `head(transactionData)` to look at the first 10 rows.

Let's check if columns we would expect to be numeric are in numeric form and date columns are in date format.

```{r Examining transcation data}
#### Examine transaction data
str(transactionData)
```

We can see that the date column is in an integer format. Let's change this to a date format.

```{r Convert DATE to date format}
#### Convert DATE column to a date format
#### A quick search online tells us that CSV and Excel integer dates begin on 30 Dec 1899
transactionData$DATE <- as.Date(transactionData$DATE, origin = "1899-12-30")
```

We should check that we are looking at the right products by examining PROD_NAME.

```{r Summary of PROD_NAME}
#### Examine PROD_NAME
summary(transactionData$PROD_NAME)
```

```{r}
transactionData %>%
  group_by(PROD_NAME) %>% 
  summarize(N = n())
```

Looks like we are definitely looking at potato chips but how can we check that these are all chips? We can do some basic text analysis by summarising the individual words in the product name.

```{r Further examine PROD_NAME}
#### Examine the words in PROD_NAME to see if there are any incorrect entries
#### such as products that are not chips
productWords <- transactionData %>% 
  distinct(PROD_NAME) %>% 
  separate_rows(PROD_NAME, sep = " ") 
names(productWords)<- "words"
```

```{r}
#### creating list of products
productInfo <- data.table(PROD_NBR = unique(transactionData$PROD_NBR),
                          PROD_NAME = unique(toupper(transactionData$PROD_NAME)))

productInfo <- productInfo %>% arrange(PROD_NBR)
```

As we are only interested in words that will tell us if the product is chips or not, let's remove all words with digits and special characters such as '&' from our set of product words. We can do this using `grepl()`.

```{r}

productWords <- as.data.table(productWords)

library(stringr)
library(stringi)

#### Removing digits
productWords[,SPECIAL := grepl("[[:digit:]]",words)]
productWords <- productWords[SPECIAL == FALSE,][,SPECIAL := NULL]

#### Removing special characters
#### Removing punctuations
productWords[,SPECIAL := grepl("[[:punct:]]",words)]

productWords <- productWords[SPECIAL == FALSE,][,SPECIAL := NULL]

#### Changing empty string to NA
productWords[words == ""] <- NA
#### Removing all empty cells
productWords <- productWords %>% 
  filter(words!="")

#### on transaction data
transactionData$PROD_NAME <- substr(transactionData$PROD_NAME, 1, nchar(transactionData$PROD_NAME))
transactionData$PROD_NAME <- gsub("\\s+"," ",transactionData$PROD_NAME )

#### on product info
productInfo$PROD_NAME <- substr(productInfo$PROD_NAME, 1, nchar(productInfo$PROD_NAME))
productInfo$PROD_NAME <- gsub("\\s+", " ", productInfo$PROD_NAME)

#### Let's look at the most common words by counting the number of times a word
#### appears and sorting them by this frequency in order of highest to lowest frequency

productWords <- data.frame(sort(table(productWords), decreasing = TRUE))
```

There are salsa products in the dataset but we are only interested in the chips category, so let's remove these.

```{r}
#### Remove salsa products

transactionData <- as.data.table(transactionData)
transactionData[, SALSA := grepl("salsa", tolower(PROD_NAME))]
transactionData[SALSA == FALSE, ][, SALSA := NULL]
transactionData <- transactionData[SALSA == FALSE,][,SALSA := NULL]

```

Next, we can use `summary()` to check summary statistics such as mean, min and max values for each feature to see if there are any obvious outliers in the data and if there are any nulls in any of the columns (`NA's : number of nulls` will appear in the output if there are any nulls).
```{r initial summary}
#### Summarise the data to check for nulls and possible outliers

summary(transactionData)
sum(is.na(transactionData))
```
There are no nulls in the columns but product quantity appears to have an outlier which we should investigate further. Let's investigate further the case where 200 packets of chips are bought in one transaction.

```{r}
#### Filter the dataset to find the outlier
outlier <- transactionData[PROD_QTY == 200,]
```

There are two transactions where 200 packets of chips are bought in one transaction and both of these transactions were by the same customer.

```{r}
#### Let's see if the customer has had other transactions
subset <- transactionData[LYLTY_CARD_NBR == 226000, ]
```

It looks like this customer has only had the two transactions over the year and is not an ordinary retail customer. The customer might be buying chips for commercial purposes instead. We'll remove this loyalty card number from further analysis.

```{r}
#### Filter out the customer based on the loyalty card number
transactionData <- transactionData %>% 
  filter(LYLTY_CARD_NBR!=226000)

#### Re-examine transaction data
summary(transactionData)

```
That’s better. Now, let’s look at the number of transaction lines over time to see if there are any obvious data issues such as missing data.
```{r}
#### Count the number of transactions by date
transactionData[, .N, by = DATE]
```
There's only 364 rows, meaning only 364 dates which indicates a missing date. Let's create a sequence of dates from 1 Jul 2018 to 30 Jun 2019 and use this to create a chart of number of transactions over time to find the missing date.
```{r fig.align ='center'}
#### Create a sequence of dates and join this the count of transactions by date 
seqofdates <- data.table(seq(as.Date('2018-07-01'), as.Date('2019-06-30'), by = "day"))
setnames(seqofdates, "date")
seqofdates$date <- as.factor(seqofdates$date)
class(seqofdates$date)

trans_by_date <- data.frame(sort(table(transactionData$DATE), decreasing = TRUE))
setnames(trans_by_date,c('date', 'freq'))
trans_by_date <- trans_by_date[order(as.Date(trans_by_date$date)),]
class(trans_by_date$date)

transaction_by_day <- merge(x = seqofdates, y = trans_by_date, by = "date", all.x = TRUE)

#### Setting plot themes to format graphs
theme_set(theme_bw())
theme_update(plot.title = element_text(hjust = 0.5))

#### Plot transactions over time
transaction_by_day$date <- as.Date(transaction_by_day$date)

ggplot(transaction_by_day, aes(x = date, y = freq)) +
 geom_line() +
 labs(x = "Day", y = "Number of transactions", title = "Transactions over time") +
 scale_x_date(breaks = "1 month") +
 theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```
We can see that there is an increase in purchases in December and a break in late December. Let's zoom in on this.
```{r fig.align = 'center'}
#### Filter to December and look at individual days
december <- 
  transaction_by_day[transaction_by_day$date >= '2018-12-01' & transaction_by_day$date <= '2018-12-31']

#### plotting transactions over december and look at individual days
ggplot(december, aes(x = date, y = freq)) +
 geom_line() +
 labs(x = "Day", y = "Number of transactions", title = "Transactions over time (December)") +
 scale_x_date(breaks = "1 day") +
 theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

```
We can see that the increase in sales occurs in the lead-up to Christmas and that there are zero sales on Christmas day itself. This is due to shops being closed on Christmas day. Now that we are satisfied that the data no longer has outliers, we can move on to creating other features such as brand of chips or pack size from PROD_NAME. We will start with pack size.

```{r Create pack size}
#### Pack size
#### We can work this out by taking the digits that are in PROD_NAME
transactionData[, PACK_SIZE := parse_number(PROD_NAME)]

#### Always check your output
#### Let's check if the pack sizes look sensible
transactionData[, .N, PACK_SIZE][order(PACK_SIZE)]
```
The largest size is 380g and the smallest size is 70g - seems sensible!
```{r}
#### Let's plot a histogram of PACK_SIZE since we know that it is a categorical variable and not a continuous variable even though it is numeric.
options(scipen = 999) #turning off scientific notation
hist(transactionData[, PACK_SIZE], main = "Histogram of the chips' packaging size",
     xlab = "Size(g)", ylab = "Total number of chips purchased", 
     col = "deepskyblue3", border = "black")
```
Pack sizes created look reasonable.

Now to create brands, we can use the first word in PROD_NAME to work out the brand name...
```{r Create brand name}
#### Brands
transactionData$BRANDS <- sapply(strsplit(transactionData$PROD_NAME, " "), `[`, 1)

# Checking brands
transactionData[, .N, by = BRANDS][order(-N)]
```
Some of the brand names look like they are of the same brands - such as RED and RRD, which are both Red Rock Deli chips. Let's combine these together.
```{r}
#### Clean brand names
transactionData[BRANDS == "Red", BRANDS := "RRD"]
transactionData[BRANDS == "Snbts", BRANDS := "Sunbites"]
transactionData[BRANDS == "Infzns", BRANDS := "Infuzions"]
transactionData[BRANDS == "WW", BRANDS := "WOOLWORTHS"]
transactionData[BRANDS == "WOOLWORTHS", BRANDS := "Woolworths"]
transactionData[BRANDS == "Smith", BRANDS := "Smiths"]
transactionData[BRANDS == "NCC", BRANDS := "NATURAL"]
transactionData[BRANDS == "Natural", BRANDS := "NATURAL"]
transactionData[BRANDS == "Dorito", BRANDS := "Doritos"]
transactionData[BRANDS == "Grain", BRANDS := "Grnwvs"]
transactionData[BRANDS == "GrnWves", BRANDS := "Grnwvs"]

#### Check again
transactionData[, .N, by = BRANDS][order(BRANDS)]
```
### Examining customer data
Now that we are happy with the transaction dataset, let's have a look at the customer dataset.
```{r 1 Exploratory data analysis}
customerData <- as.data.table(customerData)
str(customerData)
summary(customerData)
sum(is.na(customerData))

### Examining the values of lifestage and premium customers
customerData[, .N, by = LIFESTAGE][order(-N)]
customerData[, .N, by = PREMIUM_CUSTOMER][order(-N)]
```
```{r}
#### Merge transaction data to customer data
data <- merge(transactionData, customerData, all.x = TRUE)
```
As the number of rows in data is the same as that of transactionData, we can be sure that no duplicates were created. This is because we created data by setting all.x = TRUE (in other words, a left join) which means take all the rows in transactionData and find rows with matching values in shared columns and then joining the details in these rows to the x or the first mentioned table.

Let’s also check if some customers were not matched on by checking for nulls.
```{r Check for missing customer details}
colSums(is.na(data))
```
Great, there are no nulls! So all our customers in the transaction data has been accounted for in the customer dataset. Note that if you are continuing with Task 2, you may want to retain this dataset which you can write out as a csv.
```{r Code to save dataset as a csv}
write.csv(data, file = "QVI_data.csv", row.names = FALSE)
```
Data exploration is now complete!

## Data analysis on customer segments

Now that the data is ready for analysis, we can define some metrics of interest to
the client:

- Who spends the most on chips (total sales), describing customers by lifestage and how premium their general purchasing behaviour is
- How many customers are in each segment
- How many chips are bought per customer by segment
- What's the average chip price by customer segment

We could also ask our data team for more information. Examples are:

- The customer's total spend over the period and total spend for each transaction to understand what proportion of their grocery spend is on chips
- Proportion of customers in each customer egment overall to compare against the mix of customers who purchase chips 

Let's start with calculating total sales by LIFESTAGE and PREMIUM_CUSTOMER and plotting the split by these segments to describe which customer segment contribute most to chip sales.
```{r fig.width = 10, fig.align = 'center'}
#### Total sales by LIFESTAGE and PREMIUM_CUSTOMER

#total sales by life stages
sales <- data[, .(SALES = sum(TOT_SALES)),.(LIFESTAGE, PREMIUM_CUSTOMER)]

#create plot
plot <- ggplot(data = sales) + 
  geom_mosaic(aes(weight = SALES, x = product(PREMIUM_CUSTOMER, LIFESTAGE), 
                  fill = PREMIUM_CUSTOMER)) + 
  labs(x = "Lifestage", y = "Premium customer flag", 
       title = "Proportion of sales") + 
  theme(axis.text.x = element_text(angle = 50, vjust = 0.5, size = 10))

# Plot and label with proportion scales
plot + geom_text(data = ggplot_build(plot)$data[[1]], 
                 aes(x = (xmin + xmax)/2 , 
                     y = (ymin + ymax)/2, 
                     label = as.character(paste(round(.wt/sum(.wt),3)*100, '%'))))
```

There are more Mainstream - young singles/couples and Mainstream - retirees who buy chips. This contributes to there being more sales to these customer segments but this is not a major driver for the Budget - Older families segment.

Higher sales may also be driven by more units of chips being bought per customer.

Let’s have a look at this next.
```{r fig.width= 10, fig.align='center'}
#### Average number of units per customer by LIFESTAGE and PREMIUM_CUSTOMER
avg_units <- 
  data[,.(AVG = sum(PROD_QTY)/uniqueN(LYLTY_CARD_NBR)),.(LIFESTAGE, PREMIUM_CUSTOMER)][order(-AVG)]

#### create plot
ggplot(avg_units, aes(weight = AVG, x = LIFESTAGE, fill = PREMIUM_CUSTOMER)) + 
  geom_bar(position = position_dodge()) +
  labs(x = "Lifestage", y = "Average units per transaction", title = "Units per transaction") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.75, size = 7))
```
Older families and young families in general buy more chips per customer.
Let's also investigate the average price per unit chips bought for each customer segment as this is also a driver of total sales.
```{r fig.width=10, fig.align='center'}
#### average price per unit chips
avg_price <- data[,.(AVG= sum(TOT_SALES)/sum(PROD_QTY)),.(LIFESTAGE, PREMIUM_CUSTOMER)][order(-AVG)]

#### create plot
ggplot(avg_price, aes(weight=AVG, x=LIFESTAGE, fill = PREMIUM_CUSTOMER)) + 
  geom_bar(position = position_dodge())+ 
  labs(x= "Lifestage", y= "Average price per unit", title = "Price per Unit") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```
Mainstream midage and young singles and couples are more willing to pay more per packet of chips compared to their budget and premium counterparts. This may be due to premium shoppers being more likely to buy healthy snacks and when they buy chips, this is mainly for entertainment purposes rather than their own consumption. This is also supported by there being fewer premium midage and young singles and couples buying chips compared to their mainstream counterparts.

As the difference in average price per unit isn't large, we can check if this difference is statistically different.
```{r}
#### Perform an independent t-test between mainstream vs premium and budget midage and young singles and couples

price_per_unit <- data[, price := TOT_SALES/PROD_QTY]

#### create histogram of price unit to make sure the data is normally distributed

hist(price_per_unit[, price], main = "Histogram of price per unit", xlab = "Price",
     border = "black", col= "deepskyblue3")

### independent t-test between mainstream vs premium and budget midage and young singles and couples
t.test(price_per_unit[LIFESTAGE %in% c("YOUNG SINGLES/COUPLES", "MIDAGE SINGLES/COUPLES") &
                        PREMIUM_CUSTOMER == "Mainstream", price],
       price_per_unit[LIFESTAGE %in% c("YOUNG SINGLES/COUPLES", "MIDAGE SINGLES/COUPLES") &
                        PREMIUM_CUSTOMER != "Mainstream", price], alternative = "greater")

```
The t-test results in a p value of 2.2e-16, i.e. the unit price for mainstream, young and mid-age singles and couples ARE significantly higher than that of budget or premium, young and midage singles and couples.

## Deep dive into specific customer segments for insights
We have found quite a few interesting insights that we can dive deeper into.

We might want to target customer segments that contribute the most to sales to retain them or further increase sales. Let's look at Mainstream - young singles/couples. For instance, let's find out if they tend to buy a particular brand of chips.

```{r fig.align='center'}
#### Deep dive into Mainstream, young singles/couples
segment1 <- data[LIFESTAGE == "YOUNG SINGLES/COUPLES" & PREMIUM_CUSTOMER == "Mainstream",]
other <- data[!(LIFESTAGE == "YOUNG SINGLES/COUPLES" & PREMIUM_CUSTOMER == "Mainstream"),]

#### Work out of there are brands that these two customer segments prefer more than others.
####You could use a technique called affinity analysis or a-priori analysis

### brand affinity compared to the rest of the population
quantity_segment1 <- segment1[, sum(PROD_QTY)]

quantity_other <- other[, sum(PROD_QTY)]

quantity_segment1_by_brand <- segment1[, .(target_segment = sum(PROD_QTY)/quantity_segment1),
                                       by = BRANDS]

quantity_other_by_brand <- other[, .(other = sum(PROD_QTY)/quantity_other), by = BRANDS]

brand_proportions <- merge(quantity_segment1_by_brand, quantity_other_by_brand)[, affinity_to_brand := target_segment/other]

brand_proportions[order(-affinity_to_brand)]
```
```{r}
## create plot brand proportion
ggplot(brand_proportions, aes(x = reorder(BRANDS, affinity_to_brand), y= affinity_to_brand))+ 
  geom_bar(stat = "identity", fill = "pink", color = "black") +
  labs(x = "Brands", y = "Affinity to Brand", title = "Favorite Brands of Customers") + coord_flip() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```
```{r}
### calculate difference percentage of each brands

diff_percent <- brand_proportions[, .(diff_percent = (affinity_to_brand -1)*100),.(BRANDS, target_segment, other, affinity_to_brand)][order(-affinity_to_brand)]
```

We can see that:
* Around 23% mainstream young singles/couples are more likely to buy Tyrrells chips compared to the rest of the population.
* The least purchased product from mainstream young singles/couples is Burger Chips of 56% of them are less likely to buy it compared to the rest of the population.

Let’s also find out if our target segment tends to buy larger packs of chips.

```{r fig.align='center'}
# Preferred pack size compared to the rest of the population
quantity_segment1_by_pack <- segment1[, .(target_segment = (sum(PROD_QTY)/quantity_segment1)),
                                      by = PACK_SIZE]

quantity_other_by_pack <- other[, .(other = (sum(PROD_QTY)/quantity_other)), by = PACK_SIZE]

pack_proportions <- 
  merge(quantity_segment1_by_pack, quantity_other_by_pack)[, affinity_to_pack := target_segment/other]

pack_proportions[order(-affinity_to_pack)]
```
```{r}
#### create plot pack proportion
ggplot(pack_proportions, aes(x =reorder(PACK_SIZE, -affinity_to_pack), y =affinity_to_pack)) +
  geom_bar(stat = "identity", fill = "green", color = "black") +
  labs(x = "Pack Size (g)", y = "Affinity to Pack", title = "Customers Purchase by Pack Size") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```
```{r}
### calculate difference percentage of each pack size

diff_percent_pack <- pack_proportions[, .(diff_percent = (affinity_to_pack -1)*100),.(PACK_SIZE, target_segment, other, affinity_to_pack)][order(-affinity_to_pack)]
```
```{r}
#### Find the product name
data[BRANDS == "Tyrrells", unique(PROD_NAME)]
data[PACK_SIZE == 270, unique(PROD_NAME)]
```
In case of the most sales size packs, Twisties is the only brand whose products are available in 270g pack size. Therefore, mainstream young singles/couples with around 27% compared to the rest of the population tend to buy 270 g Twisties chips.

## Conclusions

Based on results we acquired from previous analyses, we can conclude that:

* The top three sales proportion are coming from Budget-Older Families, Mainstream-Young Singles/Couples, and Mainstream-Retirees shoppers. These results are reasonable because they have great numbers rather than other customers.

* Older families and young families in general buy more chips per customer.

* Mainstream midage and young singles/couples are more willing to pay more per packet of chips. These segments will become main target of promotions to increase total sales.

* Around 23% mainstream young singles/couples are more likely to buy Tyrells chips compared to the rest of the population.

* Mainstream young singles/couples with around 27% compared to the rest of the population tend to buy 270 g Twisties chips.